\textls[-0]{\thetitle}

### 2023-03-04 metatheory-condensed.tex
Changing W-type section. Need to add an example now.

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT
(and \cref{fig:fmltt-selected} \TT).
%We briefly review the base MLTT fragment first.
%
Following the current trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}
in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

@@

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT.
Together with the identity type $\TyId{\cDot}{\cDot}$, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
In the literature,
a W-type is given by a pair of types $\goodType{\Gm}{A}{}$ and $\goodType{\Gm,A}{B}{}$,
but our formulation of W-types additionally
allows a W-type signature \goodWSig{\Gm}{\wsig}{n} to
specify $n$ pairs of types (\ruleref{wsig/empty} and \ruleref{wsig/add}).
Each pair models a constructor in an inductive type,
with $\goodType{\Gm}{\wsigproj{i}{1}{\wsig}}{}$
and $\goodType{\Gm,\wsigproj{i}{1}{\wsig}}{\wsigproj{i}{2}{\wsig}}{}$
identifying the $i$-th constructor, and $\wcode{\wsig}$ giving the code of the
W-type (\ruleref{tm/w}).
%
Rule~\ruleref{tm/wsup} states that 
given two arguments $\goodTerm{\Gm}{t_1}{\wsigproj{i}{1}{\wsig}}$ and
$\goodTerm{\Gm,\sub{\wsigproj{i}{2}{\wsig}}{\SubstExt{\SubstWeak{0}}{t_1}}}{t_2}{\El{\wcode{\wsig}}}$,
we can construct a term $\Wsup{\wsig}{t_1}{t_2}$ of the W-type.

@@

>>>>

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT
(and \cref{fig:fmltt-selected} \TT).
%We briefly review the base MLTT fragment first.
%
In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}
in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

@@

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT.
Together with the identity type $\TyId{\cDot}{\cDot}$, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
Our enhanced W-type is decided by a W-type signature $\goodWSig{\Gm}{\wsig}{n}$ 
composed of a list (\ruleref{wsig/empty} and \ruleref{wsig/add}) of $n$ pairs of 
types $\{(\wsigproj{i}{1}{\wsig}, \wsigproj{i}{2}{\wsig})\}_i$,
modeling an inductive type of $n$ constructors. 
%
Rule~\ruleref{tm/wsup} states that the $i$-th pair $\goodType{\Gm}{\wsigproj{i}{1}{\wsig}}{}$ and 
$\goodType{\Gm,\wsigproj{i}{1}{\wsig}}{\wsigproj{i}{2}{\wsig}}{}$ models the $i$-th constructor---given 
two arguments $\goodTerm{\Gm}{t_1}{\wsigproj{i}{1}{\wsig}}$ and
$\goodTerm{\Gm,\sub{\wsigproj{i}{2}{\wsig}}{\SubstExt{\SubstWeak{0}}{t_1}}}{t_2}{\El{\wcode{\wsig}}}$,
we can construct a term $\Wsup{\wsig}{t_1}{t_2}$ of the W-type. $\wcode{\wsig}$ gives the code of the
W-type (\ruleref{tm/w}).

@@

\EDJ{Add a concrete example of Inductive type ``tm'' here}
